Dependently Typed Languages in Statix

### 2023-04-23 pres-thesis/presentation.tex
Update presentation.tex
Changes to presentation based on first test runs

\begin{frame}[fragile]{Background: What is Spoofax?}
	\begin{figure}
		\includegraphics[width=0.8\linewidth]{img/transformation}
	\end{figure}
\end{frame}

@@

\begin{frame}[fragile]{Background: What are Dependent Types?}
\begin{itemize}
	\item Types may depend on values!
	\begin{exampleblock}{Example}
		\texttt{concat : (A: Set) -> (n m : Nat) -> Vec A n -> Vec A m
			\\ \hspace*{48pt} -> Vec A (n + m)}
	\end{exampleblock}
% - n occurs insie a type `Vec`!
% - The type system enforces these lengths are always correct
% - Challenge: Type checking may require evaluating arbitrary terms to decide equality
\end{itemize}
\end{frame}

@@

	\begin{figure}
		\centering
		\begin{exampleblock}{Equivalent Statix code} 
			\texttt{typeOfExpr (s, Let(x, e, b)) = bt :-\\
				\,\, typeOfExpr (s, e) == et, \\
				\,\, typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt}
		\end{exampleblock}
	\end{figure}
\end{frame}

@@

	\begin{block}{What information is needed?}
		\begin{enumerate}
			\item T = Bool (Stored in Environment)
			\item x : T (Stored in Context)
		\end{enumerate}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: Environments \& Context}
	\begin{block}{How do we use scopes?}
		A scope is used as a replacement for an environment and a context.\\
		\begin{itemize}
			\item One type of edge: \verb|p|.
			\item One type of relation: \verb|NameEntry|, which can be:
				\begin{itemize}
					\item NType: Stores a type (Corresponds to a context)
					\item NSubst: Stores a substitution (Corresponds to an environment)
				\end{itemize}
		\end{itemize}
	\end{block}
\end{frame}

@@

\begin{frame}[fragile]{Type Checking: Requires Evaluation}
	\begin{figure}
		\includegraphics[width=0.9\linewidth]{screenshot001}
	\end{figure}
\end{frame}

@@

\begin{frame}[fragile]{Type Checking: Variable Capturing}
	\begin{block}{Solutions}
		\begin{enumerate}
			\item De Bruijn indices
			\item Uniquifying names
			\item Capture-avoiding substitution
			\item \textbf{Using scopes to distinguish names}
		\end{enumerate}
	\end{block}
\end{frame}

@@

\begin{frame}{On Extensibility}
	\begin{block}{How to extend the language}
		\begin{enumerate}
			\item Define the parsing rules for the new feature
			\item Create a new .stx file for the type-checking rules (and import it)
			\item Define the relations (typeOfExpr, betaReduceHead, expectBetaEq and betaReduce) in this file using the defining inference rules
		\end{enumerate}
	\end{block}
\end{frame}

\begin{frame}{Term Inference}
	\begin{example}\texttt{let id = ($\backslash$T : Type. $\backslash$x: T. x);
	\\id \_ true}
	\end{example}
	\begin{block}{How it works}
		\begin{itemize}
			\item When we encounter \texttt{expectBetaEq(e1, e2)} and \texttt{e1} or \texttt{e2} is free, we can infer it!
			\item But in Statix we can't query whether variables are free.
			\item Solution: Wrap each free constructor in \texttt{Infer}.
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}{Term Inference: Equational Unification}
	\begin{itemize}
		\item We can declare two terms to be equal if they satisfy a certain relation (such as beta equality)
		\item We do this using \emph{reduction rules}, such as:
		\begin{exampleblock}{Example of a rewrite rule}
			($\backslash$ x : T. b) a => b[x := a]
		\end{exampleblock}
	\end{itemize}
\end{frame}

\begin{frame}{Comparison with LambdaPi}
	\begin{itemize}
		\item A \emph{logical framework} in which you can embed other programming languages
		\item Supports rewrite rules
		\item Cannot define \emph{Let} because of limitations of LambdaPi
	\end{itemize}
\end{frame}

\begin{frame}{Comparison with LambdaPi: Example}
	\begin{exampleblock}{Defining FnConstruct}
		\texttt{symbol FnConstruct :
		\\\hspace{1em}Π (A : TmSort ),
		\\\hspace{1em}Π (B : TmType A → TmSort ),
		\\\hspace{1em}Π (f : Π (x : TmType A), TmType (B x)),
		\\\hspace{1em}TmType ( FnType A B);
		}

	\end{exampleblock}

	\begin{exampleblock}{Reduction rule for FnConstruct}
		\texttt{rule FnConstruct \_ \_ \$f → \$f;}
	\end{exampleblock}
\end{frame}

>>>>

@@

\begin{frame}[fragile]{Background: What is Spoofax?}
	\begin{figure}
		\includegraphics[width=0.8\linewidth]{img/transformation}
	\end{figure}
	% - used for interpreters (3+5=8) or compiler (a+b compiles to add instruction in x86)
	% - = DYNAMIC SEMANTICS
\end{frame}

\begin{frame}[fragile]{Background: What are Dependent Types?}
\begin{itemize}
	\item Types may depend on values!
	\begin{exampleblock}{Example}
		\texttt{concat : (A: Type) -> (n m : Nat) -> Vec A n -> Vec A m
			\\ \hspace*{48pt} -> Vec A (n + m)}
	\end{exampleblock}
% - n occurs insie a type `Vec`!
% - The type system enforces these lengths are always correct
% - Challenge: Type checking may require evaluating arbitrary terms to decide equality
\end{itemize}
\end{frame}

@@

	\begin{block}{Signature of \emph{typeOfExpr}}
		\texttt{typeOfExpr : scope * Expr -> Expr}
	\end{block}

	\begin{exampleblock}{Equivalent Statix code} 
		\texttt{typeOfExpr (s, Let(x, e, b)) = bt :-\\
			\,\, typeOfExpr (s, e) == et, \\
			\,\, typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt}
	\end{exampleblock}
\end{frame}

@@

	\begin{block}{What information is needed?}
		\begin{enumerate}
			\item T = Bool (Stored in Environment)
			\item b : T (Stored in Context)
		\end{enumerate}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: Environments \& Context}
	\begin{block}{How do we use scopes?}
		Scope graphs are used as a replacement for an environment and a context.
	\end{block}

	\begin{figure}
		\includegraphics[width=0.7\linewidth]{screenshot004}
	\end{figure}

@@

\begin{frame}[fragile]{Type Checking: Requires Evaluation}
	\begin{figure}
		\includegraphics[width=0.9\linewidth]{screenshot001}
	\end{figure}

	% Talk about the fact that we're now defining evaluation in Statix
\end{frame}

@@

\begin{frame}[fragile]{Type Checking: Variable Capturing}
	\begin{block}{Solutions}
		\begin{enumerate}
			\item De Bruijn indices
			\item Uniquifying names
			\item Capture-avoiding substitution
			\item \textbf{Using scopes to distinguish names}
		\end{enumerate}
	\end{block}

	\begin{block}{The type (fixed)}
		\texttt{T : Type -> T : (T, s0) -> (T, s0)}
	\end{block}
\end{frame}

@@

\begin{frame}{On Extensibility}
	\begin{block}{How to extend the language}
		\begin{enumerate}
			\item Define parsing rules
			\item Create a new .stx file
			\item Define the relations: \texttt{typeOfExpr, betaReduceHead, expectBetaEq and betaReduce}
		\end{enumerate}
	\end{block}
\end{frame}

\begin{frame}{Term Inference}
	\begin{example}\texttt{let id = ($\backslash$T : Type. $\backslash$x: T. x);
	\\id \_ true}
	\end{example}
	\begin{block}{How it works}
		\begin{itemize}
			\item When \texttt{expectBetaEq(e1, e2)} and \texttt{e1} or \texttt{e2} is free, infer!
			\item But in Statix we can't query whether variables are free.
			\item Solution: Wrap each free constructor in \texttt{Infer}.
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}{Term Inference: Equational Unification}
	\begin{itemize}
		\item We can declare two terms to be equal if they satisfy a certain relation (such as beta equality)
		\item We do this using \emph{reduction rules}, such as:
		\begin{exampleblock}{Example of a rewrite rule}
			($\backslash$ x : T. b) a => b[x := a]
		\end{exampleblock}
		\item Comparison with LambdaPi
	\end{itemize}
\end{frame}
